---
title: "lecture-06-exercises"
author: "Massimiliano Canzi"
date: "17/05/2021"
output: html_document
---

## Part 1

```{r, message = F}
library(dplyr)
```

Read the data set `iris` into R. the data set can be called with the `data()` function

```{r}
data(iris)
```

create a variable called `model` and fit a linear regression where: 

* `Petal.Width` is the outcome variable 
* `Species` is your predictor 

```{r}
iris %>% lm(...) -> model
```

Based on the regression model I am asking you to fit, what would you say is the.. (answer on the * lines)

research question? 

* 

What would you say is a reasonable hypothesis?

* 

..and the null hypothesis? 

* 

We use `lm` to combine regression with inferential statistics to 

* model the variance and the distribution of our data (through regression)
* test our hypothesis (_t_ statistic and _p_ values)

Below, display the summary of the model:

```{r}

```

Now install package `report`

```{r, eval = F}
install.packages("report")
```

```{r}
library(report)
```

Use the `report()` function on the model to learn more about your regression model

```{r}

```

What does adjusted _R squared_ suggest, in our particular case? Is the model a good fit for the data?

*

Why are there two slopes (i.e. `Speciesversicolor` and `Speciesvirginica`) but no `setosa`? 

* 

Can you plot and visually inspect the residuals?

```{r}

```

## Part 2

Now fit a similar model but adding `Petal.Length` as the second predictor of your model

```{r}
iris %>% lm(...) -> model2
```

And display the summary

```{r}

```

What would you say are the major differences between the two models?

* 

BONUS QUESTION (Optional)

We have not yet talked about interactions in linear models, but if you are familiar with what an interaction is.. 

Fit the previous model adding an interaction between Species and Petal.length. Discuss the results. 

```{r}
iris %>% lm(...) -> model3
```

* 