---
title: "Lesson 6: Multiple Linear Regression Assumptions"
subtitle: "24th/26th May, 2023"
author: "Ilaria Venagli (PhD), Dr. Massimiliano Canzi"
format:
  html:
    toc: true
    number-sections: true
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# load libraries 
library(tidyverse)
library(kableExtra)
library(unikn)
library(dlookr)
library(ggpubr)

library(lme4)
library(lmerTest)
library(sjPlot)
library(car)


# for all the chunks below, the setting "warning = F" is applied 
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

# theme set
theme_set(theme_unikn()) # set uni-konstanz theme for all plots 

# create variable "table style" and apply kableExtra styling options (see https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)
table_style <- c("striped", "hover", "condensed", "responsive") 
options(kable_styling_bootstrap_options = table_style) 

```

# Short summary

## Last time

-   `n_distinct()`: e.g., you can use it within the summarize function to count number of participants avoiding repeated rows.
-   `rowwise()`: row-wise computations - always to be combined with `ungroup()`
-   `mutate(across(c()))`
-   `mutate(c_across())`
-   `plot_likert()`
-   `log()`: log-transform continuous variables
-   `drop1(,test = "Chisq")`: model selection
-   `tab_model()`: display nice model table for knitted html/pdf documents

## Today

-   Assumptions of a linear regression model: Last week's mistake!
-   Multiple `lm` -- again! 😃
-   New functions:
    -   `report::report`
    -   `plot_model(name_model, type = "diag")`

# Homework

```{r, echo = F}
read_csv("ort.know.en.csv") %>% 
  dplyr::select(ID:write_school, 
                # reorder 
                books_en, articles_en, comics_en, posts_en, subt_en, # read en
                books_it, articles_it, comics_it, posts_it, subt_it,
                trial_number:FREQcount,
                -word_a, -word_b, -Word# read it
                ) %>% 
  rename(frequency = FREQcount)-> df.homework
```

## Exercise 1:

Create a grid with two `plot_likerts`, one should show English use at school (hence use the variables `speak_school`, `list_school`, `write_school` and `read_school`) and the other should display participants' use of EN in daily life (hence use the variables `speak_daily`, `list_daily`, `write_daily` and `read_daily`)

```{r}
### solution 

### step 1: create smaller df 
df.homework %>% select(speak_school:write_school) -> en.use.school
en.use.school %>% 
  mutate_all(~case_when(
    . == 0 ~ 0, 
    . == 0.25 ~ 1, 
    . == 0.5 ~ 2, 
    . == 0.75 ~ 3, 
    . == 1~4
  )) -> en.use.school
df.homework %>% select(speak_daily:write_daily) -> en.use.daily
en.use.daily %>% 
  mutate_all(~case_when(
    . == 0 ~ 0, 
    . == 0.25 ~ 1, 
    . == 0.5 ~ 2, 
    . == 0.75 ~ 3, 
    . == 1~4
  )) -> en.use.daily

ggarrange(
  
  # plot likert showing EN use at school 
  plot_likert(en.use.school,
            catcount = 5,
            legend.labels = c("100% IT", "25% EN, 75% IT", "50% EN, 50% IT", "75% EN, 25% IT", "100% EN"),
            title = "EN use at school", 
            axis.labels = c("Speaking", "Understanding", "Reading", "Writing"),
            geom.colors = "RdBu",
            # values = "sum.inside",
            reverse.scale = T, show.n = T), 
  
  # plot likert showing EN outside school 
  plot_likert(en.use.daily,
            catcount = 5,
            legend.labels = c("100% IT", "25% EN, 75% IT", "50% EN, 50% IT", "75% EN, 25% IT", "100% EN"),
            title = "EN use at school", 
            axis.labels = c("Speaking", "Understanding", "Reading", "Writing"),
            geom.colors = "RdBu",
            # values = "sum.inside",
            reverse.scale = T, show.n = T), 
  
  common.legend = T, 
  nrow = 2, 
  legend = "right"
  
)
```

## Exercise 2:

Create a **composite measure** of English proficiency (subjective) by computing the mean of the variables `speaking`, `reading`, `understanding` and `writing`, then create a `boxplot` showing the differences between dyslexics and non-dyslexics

```{r}
# solution 

df.homework %>% 
  # operate row-wise 
  rowwise() %>% 
  mutate(eng.prof = mean(c_across(speaking:writing))) %>% 
  ungroup() %>% 
  
  # plot 
  
  ggplot(aes(x = diagnosis, y = eng.prof, fill = diagnosis)) + geom_boxplot(alpha = .5) + ggtitle(" EN proficiency by group (TD vs. DYS) ")
  
```

# Assumptions of a linear regression

1.  **Linearity**: The relationship between each independent variable and the dependent variable is assumed to be

2.  **Independence**: The observations or data points used in the regression analysis should be independent of each other.

3.  **Constant variance**: The variance of the error terms (residuals) should be constant across all levels of the independent variables.

4.  **Normality**: The error terms (residuals) are assumed to follow a normal distribution.

5.  **No multicollinearity**: This assumption is crucial in multiple linear regression. It states that the independent variables should not be highly correlated with each other. High multicollinearity can cause instability in the estimates of the regression coefficients and affect the interpretation of the model.

### What went wrong last week?

The **df** we were using had **multiple observations for the same participants**

```{r}
read_csv("week_five.csv") -> df 
nrow(df) # 4,480 rows 
length(unique(df$ID)) # we should have 56 rows instead 
```

### IMPORTANT STEP

::: {.alert .alert-dismissible .alert-danger}
<button type="button" class="btn">

</button>

Reduce the data frame by retaining only **one observation per participant:** This has to be done here because **all the variables included in the model remain constant across all rows for each participant**. In this case, the assumption of independence would be violated.
:::

```{r}
df %>% 
  # reduce to 1 observation per participant 
  filter(trial_number == 1) %>%  # 56 
  # mutate cht to fct  
  mutate_if(is.character, as.factor) %>% 
  # scale IV 
  mutate(
    age = scale(age), age = as.double(age), 
    AoO = scale(AoO), AoO = as.double(AoO), 
    reading = scale(reading), reading = as.double(reading), 
    eng.use = scale(eng.use), eng.use = as.double(eng.use), 
    read.en = scale(read.en), read.en = as.double(read.en),
  # log transform RT 
    mean.rt = log(mean.rt)) -> df.model
```

Compare output of model with reduced data frame with output of model with bigger data frame:

## Hypothesis 1:

RT is predicted by:

-   Diagnosis
-   age
-   AoO
-   Reading exposure (not in the model now)
-   Reading proficiency
-   EN use

```{r}
# last week's model
df %>% 
  lm(formula = mean.rt ~ 
       diagnosis + 
       age + 
       AoO + 
       reading + 
       eng.use + 
       read.en
       ) -> m1 
summary(m1)

# Model with reduced data frame 
df.model %>% 
  lm(formula = mean.rt ~ 
       diagnosis + 
       age + 
       AoO + 
       reading + 
       eng.use + 
       read.en
       ) -> m1a
summary(m1a)
```

```{r}
# Check residuals of m1 and m1a 
plot_model(m1, type = "diag")
plot_model(m1a, type = "diag")
```

```{r, echo = T}
# model selection of m1a with drop1 
df.model %>% 
  lm(formula = mean.rt ~ 
       diagnosis + 
       age + 
       AoO + 
       reading + 
       eng.use +
       read.en
       ) -> m1a

## Test model 
drop1(m1a, test = "Chisq")

## Model output 
summary(m1a)
```

```{r, eval = F, echo = T}
# syntax 
plot_model(model.name, type = "diag")
vif(model.name)
```

-   `plot_model(model_name, type = "diag")`: Checks model assumptions

```{r}
# check plot model "diag"
```

-   `vif`: computes the VIF values by regressing each independent variable against the other independent variables in the model

```{r}
# check vif 

```

-   **VIF = 1:** indicates no multicollinearity, meaning there is no correlation between the independent variable and other independent variables in the model

-   **VIF** **between 1 and 5** are generally considered acceptable, indicating low to moderate multicollinearity

-   **VIF values above 5 or 10** (depending on the field) are often considered high, suggesting the presence of multicollinearity.

```{r}
# plot model 
```

### Report model

```{r, eval = F}
# syntax 
report::report(model.name)
```

```{r}
# report 
```

## Hypothesis 2

```{r}
df.model %>% 
  lm(formula = accuracy_tot ~ 
       age + 
       AoO + 
       reading + 
       diagnosis + 
       eng.use + 
       read.en
     ) -> m2

## Model selection
#drop1(m2, test = "Chisq")

## Model output
#summary(m2)
```

```{r}
# check assumptions 
```

```{r}
# plot model 
```

```{r}
# report model
```
